---
title: "Devon's_Playaround_space"
output: html_document
date: "2025-11-13"
---
---- Specific to transplants -----

```{r}
 library(haven)
library(tidyverse)
library(scales)
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)
library(plotly)

```


```{r}
liver_data_raw = read_sas("/Users/devonpark/Library/CloudStorage/OneDrive-ColumbiaUniversityIrvingMedicalCenter/tx_li.sas7bdat")

```

GENERAL:

```{r}
liver_data_1 = liver_data_raw |> 
  mutate(tx_date = as.Date(REC_TX_DT),
        listing_date = as.Date(CAN_LISTING_DT),
        tx_year = year(tx_date),
        days_listed = as.numeric(difftime(tx_date, listing_date, units = "days")),
        alcohol_subtype = case_when(
                CAN_DGN %in% c(4215, 4216) | CAN_DGN2 %in% c(4215, 4216) ~ "Alcohol-related Cirrhosis",
                CAN_DGN == 4217 | CAN_DGN2 == 4217 ~ "Acute Alcohol-related Hepatitis",
                TRUE ~ "Other")) %>% # days on waitlist
filter(!is.na(PERS_ID) & !is.na(tx_date) & tx_year >= 2002 & REC_AGE_AT_TX >= 18,
      alcohol_subtype != "Other") |> # MELD era only, Adults only
arrange(PERS_ID, tx_date) |> 
  janitor::clean_names()



liver_data_clean_1 = 
  liver_data_1 |>
    mutate(
      don_race_srtr_std = don_race_srtr,
      don_race_srtr_std = case_when(
        don_race_srtr_std == "ASIAN"   ~ "Asian",
        don_race_srtr_std == "WHITE"   ~ "White",
        don_race_srtr_std == "BLACK"   ~ "Black or African American",
        don_race_srtr_std == "NATIVE"  ~ "American Indian or Alaska Native",
        don_race_srtr_std == "MULTI"   ~ "Multi-Racial",
        don_race_srtr_std == "PACIFIC" ~ "Native Hawaiian or Other Pacific Islander",
        don_race_srtr_std == ""        ~ "NOT SPECIFIED",
        TRUE ~ don_race_srtr_std #For all rows that did not match any of the previous recoding conditions, keep the original value
      )
    )
```


```{r}
liver_data_clean_1 |>
  count(tx_year, don_race_srtr_std) |>
  group_by(tx_year) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = tx_year, y = pct, fill = don_race_srtr_std)) +
  geom_col(color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Donor Race by Transplant Year",
    x = "Transplant Year",
    y = "Percent",
    fill = "Donor Race"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

-------------------------------------------------------------------

```{r}
liver_data = liver_data_raw |> 
  mutate(tx_date = as.Date(REC_TX_DT),
        listing_date = as.Date(CAN_LISTING_DT),
        tx_year = year(tx_date),
        days_listed = as.numeric(difftime(tx_date, listing_date, units = "days")),
        alcohol_subtype = case_when(
                CAN_DGN %in% c(4215, 4216) | CAN_DGN2 %in% c(4215, 4216) ~ "Alcohol-related Cirrhosis",
                CAN_DGN == 4217 | CAN_DGN2 == 4217 ~ "Acute Alcohol-related Hepatitis",
                TRUE ~ "Other")) %>% # days on waitlist
filter(!is.na(PERS_ID) & !is.na(tx_date) & tx_year >= 2020 & REC_AGE_AT_TX >= 18,
      alcohol_subtype != "Other") |> # MELD era only, Adults only
arrange(PERS_ID, tx_date) |> 
  janitor::clean_names()
```

Looking at race:
```{r}
#Clean don_Race column

liver_data_clean = 
liver_data |> 
  mutate(
    
    don_race_raw = don_race,#Clean original race codes
    true_missing = is.na(don_race_raw) | don_race_raw %in% c("NA", "", " "), # 2. Identify TRUE missing before numeric conversion
   
    don_race_num = suppressWarnings(as.numeric(trimws(as.character(don_race_raw)))),  #Converts to numeric (will produce NA for non-numeric multi-racial codes)

    don_race = case_when(
      don_race_num == 8 ~ "White",
      don_race_num == 16 ~ "Black or African American",
      don_race_num == 32 ~ "American Indian or Alaska Native",
      don_race_num == 64 ~ "Asian",
      don_race_num == 128 ~ "Native Hawaiian or Other Pacific Islander",
      don_race_num == 256 ~ "Arab or Middle Eastern",
      don_race_num == 512 ~ "Indian Sub-continent",
      don_race_num == 1024 ~ "Unknown (for Donor Referral only)",
      don_race_num == 2000 ~ "Hispanic/Latino",

      don_race_num %in% c(24, 40, 48, 56, 72, 80, 88, 96, 104, 112, 120, 
                          136, 144, 192, 200, 232) ~ "Multi-Racial",

      true_missing ~ "Missing", #see if there are TRUE missing BEFORE numeric conversion

      is.na(don_race_num) ~ "Multi-Racial", #Numeric conversion failures → Multi-Racial (your new rule)

      TRUE ~ as.character(don_race_raw) #Fallback in case
    )
  ) 
  
```


```{r}
#liver_data_clean |> 
#  distinct(don_race_srtr)

liver_data_clean = 
  liver_data |>
    mutate(
      don_ty_factor = case_when(
      don_ty == "C" ~ "Deceased Donor",
      don_ty == "L" ~ "Living Donor",
      TRUE          ~ "Other / Unknown"),
    don_ty_factor = factor(don_ty_factor, levels = c("Deceased Donor", "Living Donor", "Other / Unknown")),
    
      don_race_srtr_std = don_race_srtr,
      don_race_srtr_std = case_when(
        don_race_srtr_std == "ASIAN"   ~ "Asian",
        don_race_srtr_std == "WHITE"   ~ "White",
        don_race_srtr_std == "BLACK"   ~ "Black or African American",
        don_race_srtr_std == "NATIVE"  ~ "American Indian or Alaska Native",
        don_race_srtr_std == "MULTI"   ~ "Multi-Racial",
        don_race_srtr_std == "PACIFIC" ~ "Native Hawaiian or Other Pacific Islander",
        don_race_srtr_std == ""        ~ "NOT SPECIFIED",
        TRUE ~ don_race_srtr_std #For all rows that did not match any of the previous recoding conditions, keep the original value
      )
    )
  

```

Race breakdown since 2020 annually for all donors 
```{r}
liver_data_clean |>
  count(tx_year, don_race_srtr_std) |>
  group_by(tx_year) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = tx_year, y = pct, fill = don_race_srtr_std)) +
  geom_col(color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Donor Race by Transplant Year",
    x = "Transplant Year",
    y = "Percent",
    fill = "Donor Race"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


Side by side Deceased vs Living Donor Race Break Down (combining years 2020-2025)
```{r}
interactive_race = liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>
  count(don_ty_factor, don_race_srtr_std) |>
  group_by(don_ty_factor) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = don_ty_factor, 
             y = pct, 
             fill = don_race_srtr_std,
             text = paste0(
      "Donor Type: ", don_ty_factor, "<br>",
      "Race: ", don_race_srtr_std, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n
    ))) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Race Composition by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Race"
  )+
  theme_minimal()
```

Interactive Version:
```{r}
ggplotly(interactive_race, tooltip = "text")
```


Exploratory: Look at deceased vs living donors and gender

```{r}
liver_data_clean |> count(don_ty)
liver_data_clean |> count(don_gender)
liver_data_clean |> filter(don_ty == "L" | don_ty == "C") |> count(don_gender)
liver_data_clean |> filter(don_ty == "C") |> count(don_gender)
liver_data_clean |> filter(don_ty == "L") |> count(don_gender)

```

Exploratory: Look at citizenship
```{r}
liver_data_clean |> count(don_citizenship)
#916 NA values

#Code	Description
#1	1: US Citizen
#2	2: Resident Alien
#3	3: Non-Resident Alien, Specify Country
#4	4: Non-US Citizen/US Resident
#5	5: Non-US Citizen/Non-US Resident
#998	998: Unknown
```

```{r}

liver_data_clean <- liver_data_clean |>
  mutate(
    don_citizenship_label = case_when(
      don_citizenship == 1 ~ "US Citizen",
      don_citizenship == 2 ~ "Resident Alien",
      don_citizenship == 3 ~ "Non-Resident Alien (Specify)",
      don_citizenship == 4 ~ "Non-US Citizen/US Resident",
      don_citizenship == 5 ~ "Non-US Citizen/Non-US Resident",
      don_citizenship == 998 ~ "Unknown",
      is.na(don_citizenship) ~ "Missing",
      TRUE ~ "Other"
    ),
    don_ty_label = case_when(
      don_ty == "C" ~ "Deceased Donor",
      don_ty == "L" ~ "Living Donor",
      TRUE          ~ "Other / Unknown"
    )
  )
```

Stacked barchart of citizenship: Decesased v Living Donor
```{r}

liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>   # only deceased + living
  count(don_ty_label, don_citizenship_label) |>
  group_by(don_ty_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = don_ty_label, y = pct, fill = don_citizenship_label)) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Citizenship Status by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Citizenship"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11),
    legend.position = "right"
  )
```

Interactive plotly version
```{r}

plotly_citizenship = liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>   # only deceased + living
  count(don_ty_label, don_citizenship_label) |>
  group_by(don_ty_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(
    x = don_ty_label,
    y = pct,
    fill = don_citizenship_label,
    text = paste0(
      "Donor type: ", don_ty_label, "<br>",
      "Citizenship: ", don_citizenship_label, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n
    )
  )) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Citizenship Status by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Citizenship"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11),
    legend.position = "right"
  )


ggplotly(plotly_citizenship, tooltip = "text")
```





-------------------------------
OTHER:


```{r}
# Visualise don_race

liver_data_clean |>
  count(don_race, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = n, y = reorder(don_race, n))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(pct, 3), "%")), 
            hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Donor Race",
    x = "Count",
    y = "Race"
  ) +
  theme_minimal() +
  xlim(0, max(liver_data_clean |> count(don_race) |> pull(n)) * 1.15)
```

Check for mismatches between don_race and don_race_srtr (but actually don_Race_srtr_std)
```{r}
#FILTER OUT ROWS THAT ARE counted as "Unknown (for Donor Referral only)" in don_race
mismatched_race_x = 
  liver_data_clean |> 
  select(pers_id, don_race, don_race_srtr_std) |>
  filter(don_race != "Unknown (for Donor Referral only)") |> 
  mutate(
    race_match = don_race == don_race_srtr_std #creates a logical vector
  ) |> 
  filter(race_match == FALSE)


mismatched_race_y = 
  liver_data_clean |> 
  select(pers_id, don_race, don_race_srtr_std) |>
  filter(don_race == "Unknown (for Donor Referral only)") |> 
  mutate(
    race_match = don_race == don_race_srtr_std #creates a logical vector
  ) |> 
  filter(race_match == FALSE)

```

Plot for DECEASED don_Race_srtr_std:
```{r}
liver_data_clean |>
  filter(don_ty == "L") |> 
  count(don_race_srtr_std, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = n, y = reorder(don_race_srtr_std, n))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(pct, 2), "%")), 
            hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Living Donor Race",
    x = "Count",
    y = "Race"
  ) +
  theme_minimal() +
  xlim(0, 10000)
```


Plot for DECEASED don_Race_srtr_std:
```{r}
liver_data_clean |>
  filter(don_ty == "C") |> 
  count(don_race_srtr_std, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = n, y = reorder(don_race_srtr_std, n))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(pct, 2), "%")), 
            hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Deceased Donor Race",
    x = "Count",
    y = "Race"
  ) +
  theme_minimal() +
  xlim(0, 160000)
```

Plot for don_Race_srtr_std:
```{r}
liver_data_clean |>
  count(don_race_srtr_std, sort = TRUE) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = n, y = reorder(don_race_srtr_std, n))) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = paste0(round(pct, 3), "%")), 
            hjust = -0.1, size = 3.5) +
  labs(
    title = "Distribution of Donor Race",
    x = "Count",
    y = "Race"
  ) +
  theme_minimal() +
  xlim(0, max(liver_data_clean |> count(don_race_srtr_std) |> pull(n)) * 1.15)
```

Bar chart of citizenship
```{r}

liver_data_clean |>
count(don_citizenship_label) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = pct, y = fct_reorder(don_citizenship_label, pct))) +
  geom_col() +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            hjust = -0.1, size = 3.3) +
  scale_x_continuous(labels = percent_format(), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Donor Citizenship Status",
    x = "Percent of Donors",
    y = "Citizenship"
  )
```

Donor cause of death:
```{r}

liver_data_clean |> count(don_cad_don_cod)

#7120 NA values
#Code	Description
#1	1: ANOXIA
#2	2: CEREBROVASCULAR/STROKE
#3	3: HEAD TRAUMA
#4	4: CNS TUMOR
#998	998: UNKNOWN
#999	999: OTHER SPECIFY
```


Donor death Circumstance 
```{r}

liver_data_clean |> count(don_death_mech)

#Code	Description
#1	1: MVA
#2	2: SUICIDE
#3	3: HOMICIDE
#4	4: CHILD-ABUSE
#5	5: NON-MVA
#6	6: DEATH FROM NATURAL CAUSES
#997	997: NONE OF THE ABOVE
#998	998: UNKNOWN
```


Donor death Mechanism
```{r}

liver_data_clean |> count(don_death_circum)

#Code	Description
#1	1: DROWNING
#2	2: SEIZURE
#3	3: DRUG INTOXICATION
#4	4: ASPHYXIATION
#5	5: CARDIOVASCULAR
#6	6: ELECTRICAL
#7	7: GUNSHOT WOUND
#8	8: STAB
#9	9: BLUNT INJURY
#10	10: SIDS
#11	11: INTRACRANIAL HEMORRHAGE/STROKE
#12	12: DEATH FROM NATURAL CAUSES
#995	995: 995-Gunshot/stab wound (Pre-OTIS)
#997	997: NONE OF THE ABOVE
#998	998: UNKNOWN
```





